---
title: "Investigating traffic patterns of public transportation in NYC influenced by different conditions"
author: "Advait Ashtikar"
format: pdf
editor: visual
---

### Loading libraries and reading in data for subways and buses

```{r}
library(dplyr)
library(readr)
library(lubridate)
library(readxl)
library(ggplot2)
library(tidyr)
library(scales)
library(prophet)
```

```{r}
url1 <- "MTA_Daily_Ridership_Data__Beginning_2020.csv"
df <- read_csv(url1)

head(df)
```

### Creating variable specific data frames

```{r}
# subways data frame 
sub_df <-
  df %>%
  select('Date', 'Subways: Total Estimated Ridership', 'Subways: % of Comparable Pre-Pandemic Day') %>%
  na.omit %>%
  filter(!grepl("2023$", Date),
         !grepl("2020$", Date)) %>%
  mutate("Date" = mdy(Date),
         "Day of Week" = weekdays(Date)) %>%
  select('Day of Week', 'Date', 'Subways: Total Estimated Ridership', 'Subways: % of Comparable Pre-Pandemic Day')
sub_df
```

```{r}
# buses data frame
bus_df <-
  df %>%
  select('Date', 'Buses: Total Estimated Ridership', 'Buses: % of Comparable Pre-Pandemic Day') %>%
  na.omit() %>%
  filter(!grepl("2023$", Date),
         !grepl("2020$", Date)) %>%
  mutate("Date" = mdy(Date),
         "Day of Week" = weekdays(Date)) %>%
  select('Day of Week', 'Date', 'Buses: Total Estimated Ridership', 'Buses: % of Comparable Pre-Pandemic Day')
bus_df

```

### Reading in data for weather

```{r}
url2 <- "weather_nyc_2021_2022.xlsx"
weather_df <- read_excel(url2)


head(weather_df)
```

### Data wrangling weather data frame

```{r}
weather_df <-
  weather_df %>%
  select('datetime', 'tempmax', 'tempmin', 'temp', 'precip', 'snow', 'snowdepth', 'windspeed', 'conditions', 'icon')
weather_df

```

### Joining the weather data frame with both the bus and subway data frame

```{r}
sub_df <-
  sub_df %>%
  full_join(weather_df, by = c("Date" = "datetime"))
sub_df
```

```{r}
bus_df <-
  bus_df %>%
  full_join(weather_df, by = c("Date" = "datetime"))
bus_df
```

### EDA

Summary statistics for the datasets

```{r}
summary(sub_df)
summary(bus_df)
```

Correlation between variables

```{r}
cor(sub_df %>%
      select(-c('Day of Week', 'Date', 'conditions', 'icon')))
cor(bus_df %>%
      select(-c('Day of Week', 'Date', 'conditions', 'icon')))
```

Visualizations:

1.  Ridership trends over time

```{r}
ggplot() +
  geom_line(data = sub_df, aes(x = Date, y = `Subways: Total Estimated Ridership`, color = "Subway Ridership")) +
  geom_line(data = bus_df, aes(x = Date, y = `Buses: Total Estimated Ridership`, color = "Bus Ridership")) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Daily Ridership for Subways and Buses",
       x = "Date",
       y = "Total Estimated Ridership",
       color = "Transportation") +
  theme_minimal()

```

2.  Ridership by day of the week

```{r}
sub_df_agg <- sub_df %>% group_by(`Day of Week`) %>% summarise(`Subways: Total Estimated Ridership` = mean(`Subways: Total Estimated Ridership`))
bus_df_agg <- bus_df %>% group_by(`Day of Week`) %>% summarise(`Buses: Total Estimated Ridership` = mean(`Buses: Total Estimated Ridership`))

ridership_agg <- full_join(sub_df_agg, bus_df_agg, by = "Day of Week") %>% gather(key = "Transportation", value = "Ridership", -`Day of Week`)

ggplot(data = ridership_agg, aes(x = `Day of Week`, y = Ridership, fill = Transportation)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Average Ridership by Day of the Week",
       x = "Day of the Week",
       y = "Average Ridership",
       fill = "Transportation") +
  theme_minimal()

```

3.  Scatterplot of ridership against temperature

```{r}
ggplot(sub_df, aes(x = temp, y = `Subways: Total Estimated Ridership`, color = `Subways: % of Comparable Pre-Pandemic Day`)) +
  geom_point() +
  scale_color_gradient(low = "red", high = "blue") +
  labs(title = "Subway Ridership vs. Temperature",
       x = "Temperature (°F)",
       y = "Total Estimated Ridership",
       color = "Percentage of\nPre-Pandemic Ridership") +
  theme_minimal()
```

```{r}
ggplot(bus_df, aes(x = temp, y = `Buses: Total Estimated Ridership`, color = `Buses: % of Comparable Pre-Pandemic Day`)) +
  geom_point() +
  scale_color_gradient(low = "red", high = "blue") +
  labs(title = "Bus Ridership vs. Temperature",
       x = "Temperature (°F)",
       y = "Total Estimated Ridership",
       color = "Percentage of\nPre-Pandemic Ridership") +
  theme_minimal()
```

4.  Boxplot of ridership by day of the week

```{r}
ggplot(sub_df, aes(x = `Day of Week`, y = `Subways: Total Estimated Ridership`, fill = `Day of Week`)) +
  geom_boxplot() +
  labs(title = "Subway Ridership by Day of the Week",
       x = "Day of the Week",
       y = "Total Estimated Ridership") +
  theme_minimal()
```

```{r}
ggplot(bus_df, aes(x = `Day of Week`, y = `Buses: Total Estimated Ridership`, fill = `Day of Week`)) +
  geom_boxplot() +
  labs(title = "Bus Ridership by Day of the Week",
       x = "Day of the Week",
       y = "Total Estimated Ridership") +
  theme_minimal()

```

5.  Ridership vs Precipitation

```{r}
ggplot(sub_df, aes(x = precip, y = `Subways: Total Estimated Ridership`, color = `Subways: % of Comparable Pre-Pandemic Day`)) +
  geom_point() +
  scale_color_gradient(low = "red", high = "blue") +
  labs(title = "Subway Ridership vs. Precipitation",
       x = "Precipitation (inches)",
       y = "Total Estimated Ridership",
       color = "Percentage of\nPre-Pandemic Ridership") +
  theme_minimal()
```

```{r}
ggplot(bus_df, aes(x = precip, y = `Buses: Total Estimated Ridership`, color = `Buses: % of Comparable Pre-Pandemic Day`)) +
  geom_point() +
  scale_color_gradient(low = "red", high = "blue") +
  labs(title = "Bus Ridership vs. Precipitation",
       x = "Precipitation (inches)",
       y = "Total Estimated Ridership",
       color = "Percentage of\nPre-Pandemic Ridership") +
  theme_minimal()
```

6.  Histogram of ridership distribution

```{r}
ggplot(sub_df, aes(x = `Subways: Total Estimated Ridership`)) +
  geom_histogram(binwidth = 50000, fill = "steelblue", color = "black") +
  labs(title = "Distribution of Subway Ridership",
       x = "Total Estimated Ridership",
       y = "Frequency") +
  theme_minimal()
```

```{r}
ggplot(bus_df, aes(x = `Buses: Total Estimated Ridership`)) +
  geom_histogram(binwidth = 10000, fill = "steelblue", color = "black") +
  labs(title = "Distribution of Bus Ridership",
       x = "Total Estimated Ridership",
       y = "Frequency") +
  theme_minimal()
```

### Machine Learning

Adding dataset with holiday information

```{r}
holiday_df <- data.frame(
  holiday = c("New Year's Day", "Independence Day", "Thanksgiving", "Christmas"),
  ds = as.Date(c("2021-01-01", "2021-07-04", "2021-11-25", "2021-12-25")),
  lower_window = 0,
  upper_window = 1
)

holiday_df %>% head

#Merging the dataset
sub_df <- sub_df %>% mutate(isholiday = ifelse(Date %in% holiday_df$ds, 1, 0))
bus_df <- bus_df %>% mutate(isholiday = ifelse(Date %in% holiday_df$ds, 1, 0))
```

Creating the train/test datasets

```{r}
# Split the data into training and testing sets
train_sub_df <- sub_df %>% filter(Date < as.Date("2022-01-01"))
test_sub_df <- sub_df %>% filter(Date >= as.Date("2022-01-01"))

train_bus_df <- bus_df %>% filter(Date < as.Date("2022-01-01"))
test_bus_df <- bus_df %>% filter(Date >= as.Date("2022-01-01"))
```

Preparing the model to use the `prophet` function

```{r}
# Create a function to prepare the data for the Prophet model
prepare_data_for_prophet <- function(df, y_column) {
  df <- df %>%
    select(Date, y = !!y_column) %>%
    rename(ds = Date)
  return(df)
}

# Prepare the data for the Prophet model
train_sub_prophet <- prepare_data_for_prophet(train_sub_df, 'Subways: Total Estimated Ridership')
test_sub_prophet <- prepare_data_for_prophet(test_sub_df, 'Subways: Total Estimated Ridership')
  
train_bus_prophet <- prepare_data_for_prophet(train_bus_df, 'Buses: Total Estimated Ridership')
test_bus_prophet <- prepare_data_for_prophet(test_bus_df, 'Buses: Total Estimated Ridership')
```

Using the prophet model to forecast ridership

```{r}
sub_prophet_model <- prophet(df = train_sub_prophet, holidays = holiday_df, yearly.seasonality = TRUE)
sub_forecast <- predict(sub_prophet_model, test_sub_prophet)

bus_prophet_model <- prophet(df = train_bus_prophet, holidays = holiday_df, yearly.seasonality = TRUE)
bus_forecast <- predict(bus_prophet_model, test_bus_prophet)
```

Combining actual and forecasted values for comparison

```{r}
sub_comparison <- test_sub_prophet %>%
  left_join(sub_forecast %>% select(ds, yhat), by = c("ds")) %>%
  rename(actual = y, forecast = yhat)

bus_comparison <- test_bus_prophet %>%
  left_join(bus_forecast %>% select(ds, yhat), by = c("ds")) %>%
  rename(actual = y, forecast = yhat)
```

Calculating MAE, MSE and RMSE

```{r}
calculate_metrics <- function(actual, forecast){
  mae <- mean(abs(actual - forecast))
  mse <- mean((actual - forecast)^2)
  rmse <- sqrt(mse)
  
  return(list(MAE = mae, MSE = mse, RMSE = rmse))
}

sub_metrics <- calculate_metrics(sub_comparison$actual, sub_comparison$forecast)
sub_metrics
```

```{r}
bus_metrics <- calculate_metrics(bus_comparison$actual, bus_comparison$forecast)
bus_metrics
```

### Inferences from ML

Plotting actual vs forecasted values

```{r}
ggplot(sub_comparison, aes(x = ds)) +
  geom_line(aes(y = actual, color = "Actual")) +
  geom_line(aes(y = forecast, color = "Forecast")) +
  labs(title = "Subway Ridership: Actual vs Forecast",
       x = "Date",
       y = "Ridership",
       color = "Legend") +
  theme_minimal()
```

```{r}
ggplot(bus_comparison, aes(x = ds)) +
  geom_line(aes(y = actual, color = "Actual")) +
  geom_line(aes(y = forecast, color = "Forecast")) +
  labs(title = "Bus Ridership: Actual vs Forecast",
       x = "Date",
       y = "Ridership",
       color = "Legend") +
  theme_minimal()
```

Visualizing Residuals

```{r}
sub_comparison <- sub_comparison %>% mutate(residuals = actual - forecast)
bus_comparison <- bus_comparison %>% mutate(residuals = actual - forecast)

ggplot(sub_comparison, aes(x = ds, y = residuals)) +
  geom_point(color = "steelblue") +
  geom_smooth(se = FALSE, color = "red", linetype = "dashed") +
  labs(title = "Subway Ridership Residuals",
       x = "Date",
       y = "Residuals") +
  theme_minimal()
```

```{r}
ggplot(bus_comparison, aes(x = ds, y = residuals)) +
  geom_point(color = "steelblue") +
  geom_smooth(se = FALSE, color = "red", linetype = "dashed") +
  labs(title = "Bus Ridership Residuals",
       x = "Date",
       y = "Residuals") +
  theme_minimal()
```
